An Answer Set Programming Framework for Reasoning about Agents' Beliefs and Truthfulness of Statements

### 2020-07-05 main.tex
Typesetting fixed.

% Title and Author Information Must Immediately Follow
% the pdfinfo within the preamble
%
\title{An Answer Set Programming Framework for Reasoning about Agents' Beliefs and Truthfulness of Statements} %\\
\author{Marcello Balduccini$^1$ \And 
Michael Gelfond$^2$ \And
Enrico Pontelli$^3$ \and Tran Cao Son$^3$ 
\affiliations
$^1$Saint Joseph's University \and 
$^2$Texas Tech University\and 
$^3$New Mexico State University
\emails
   mbalducc@sju.edu,   
   michael.gelfond@ttu.edu,
\{epontell, tson\}@cs.nmsu.edu 
} %\\

@@

\begin{lstlisting}[language=clingo,caption=Set $RD^s$ of statements for the self-defense example, label=lst:pi_S, mathescape=true,xleftmargin=.01\textwidth, breaklines=true]
causes(appears(X),is(X,present),[]).
default($d^s_1$,inImminentDanger,[has(att,k),
  is(att,present),is(att,menc)])
default($d^s_2$,$\neg$inImminentDanger,[has(att,k),
  is(att,present),is(att,menc),is(k,fake)])
prefer($d^s_2$,$d^s_1$, [ ]).
default($d^s_3$,$\neg$inImminentDanger,
  [at(a1), is(a1,good)])
prefer($d^s_1$,$d^s_3$, [ ]).
\end{lstlisting}
The representation uses fluents $has(p,o)$ (p has/holds object o), $is(o,a)$ (object o has attribute a), and $at(l)$ (Bob is at location $l$). The objects are $k$ (the knife), $a_1$ (the good area of the city), and $att$ (the attacker). Additionally, we attributes constants $present$ (for attribute ``physically present''), $fake$ (attribute of the knife), $menc$ (menacing-looking), and $good$ (attribute of the area of the city). The actions are $appears(p)$ (person $p$ appears) and $shoots(p)$ (Bob shoots $p$). The effect of the former is described by the $causes$ statement; the latter action has no effects within the scope of the example. The intuitive meaning of the defaults is:
\begin{itemize}
\item $d^s_1$: If there is someone who is menacing-looking and holding a knife, one will normally believe to be in imminent danger
\item $d^s_2$: In the above case, if the knife is fake, one will not normally believe to be in imminent danger
\item $d^s_3$: One does not normally believe to be in imminent danger in a good area of the city
\end{itemize} 
As can be seen from the $prefer$ statements, $d^s_2$ is the most preferred default and $d^s_3$ the least preferred.
The initial part of the story is captured by:
\[
\begin{array}{lll}
O^s_{init}  =  \left\{
\begin{array}{ll} 
obs(at(a_1),0).   %\ \ \ \ 
& obs(is(a_1,good),0). \\
obs(\neg is(att,present),0). & 
occ(appears(att),0). \\    % the guy appears
obs(is(att,menc),1). &  %\ \ \ \     % the guy is menacing-looking
obs(has(att,k),1).
\end{array}
\right\}
\end{array}
\]
Let us consider what our framework yields for $KB^s_{init}=\langle O^s_{init},RD^s \rangle$. One can check that the answer set of $\Pi(KB^s_{init})$ contains $h(at(a_1),0)$ and $h(is(a_1,good),0)$. The only applicable default is $d^s_3$, yielding $believes(\neg inImminentDanger,0)$, i.e. Bob does not believe to be in imminent danger at time step $0$. The two possible evolutions of the story are formalized as follows:
\begin{itemize}
\item Scenario 1: \[
\begin{array}{lll}
O^s_{1} & = & \left\{
\begin{array}{l} 
occ(shoots(bob,att),1). 
\end{array}
\right\}\\
\ \\
\multicolumn{3}{l}{KB^s_1=\langle O^S_{init} \cup O^s_1,RD^s \rangle}
\end{array}
\]
The time step of interest is $1$. Note that $h(is(att,menc),1)$ and $h(has(att,k),1)$ follow trivially from the observations. Defaults $d^s_1$ and $d^s_3$ are thus applicable, while $d^{s}_2$ is not.  Because $d^s_1$ is preferred to $d^s_3$, the answer set of $\Pi(KB^s_1)$ contains $believes(inImminentDanger, 1)$, meaning that, at that time Bob pulled the trigger, he believed to be in imminent danger at the time of the shooting. It is worth noting how the framework is able to capture the evolution of Bob's beliefs over time.  
\item Scenario 2: \[
\begin{array}{lll}
O^s_{2} & = & \left\{
\begin{array}{l} 
obs(is(k,fake),1). \\   % the knife is fake
occ(shoots(bob,att),1).
\end{array}
\right\} \\
\ \\
\multicolumn{3}{l}{KB^s_2=\langle O^S_{init} \cup O^s_2,RD^s \rangle}
\end{array}
\]
At time step $1$, Bob has observed that the knife is fake and, thus, $h(is(k,fake),1)$ holds. Hence, all defaults are applicable.  Because $d^s_2$ is preferred to $d^s_1$, the answer set of $\Pi(KB^s_2)$ contains $believes(\neg inImminentDanger, 1)$. That is, Bob did not initially believe to be in imminent danger and did not believe so at the time of the shooting either. 
\end{itemize}

@@

%
\begin{figwindow}[3,r,%
        {\includegraphics[width=.5\textwidth]{mim}},%
        {MITM attack carried out by Stuxnet: safe commands sent by the PLC (green) are replaced by dangerous ones (red); alarm readings from the centrifuge's sensors (red) are replaced by  safe ones\label{fig:stuxnet}}]
\end{figwindow}
The behavior of Stuxnet's MITM component is outlined in Figure \ref{fig:stuxnet}. Its component operates by intercepting the commands sent by a PLC to a connected centrifuge. The malware first increased the speed of the centrifuge above normal levels for a short amount of time, and later slowed it down below normal levels for a longer period of time. It is believed that the resulting stress caused components of the centrifuge to expand and eventually destroy it. Under normal conditions, sensors installed in the centrifuge would have alerted the PLC---and its users---about the abnormal conditions, giving them a chance to shut down the system before damage occurred. However, as part of the MITM attack, Stuxnet also intercepted the sensor readings from the centrifuge, and sent to the PLC fake readings based on previous recordings that indicated that the system was operating normally (this is known as a \emph{replay attack}).

@@

The relevant information can be formalized as shown below. The signature is defined so that $alert$ and $winter$ are fluents, while all others are beliefs. %using the following statements:
\iffalse
{
\[
\begin{array}{l}
default(d_1^m, hot\_room, [alert]). \\
default(d_2^m, \neg hot\_room, [cold\_weather]). \\
prefer(d_2^m,d_1^m, [ \: ]).\\
rule(r_1^m,cold\_weather,[winter]). \\
default(d_3^m,overheat,[alert, \neg hot\_room]).
\end{array}
\]
}
\fi
%
\begin{lstlisting}[language=clingo,caption=Relevant Information, label=lst:pi_M, mathescape=true,xleftmargin=.01\textwidth, breaklines=true]
default($d_1^m$, hot_room, [alert]). 
default($d_2^m$, $\neg$hot_room, [cold_weather]). 
prefer($d_2^m$,$d_1^m$, [ ]). 
rule($r_1^m$,cold_weather,[winter]). 
default($d_3^m$,overheat,[alert, $\neg$hot_room]).
\end{lstlisting}

@@

\section*{Acknowledgment}

>>>>

% Title and Author Information Must Immediately Follow
% the pdfinfo within the preamble
%
\title{An Answer Set Programming Framework for Reasoning about Agents' Beliefs and Truthfulness of Statements} %\\
\author{Marcello Balduccini$^1$ \and 
Michael Gelfond$^2$ \and
Enrico Pontelli$^3$ \and 
Tran Cao Son$^3$ 
\affiliations
$^1$Saint Joseph's University \\ 
$^2$Texas Tech University\\ 
$^3$New Mexico State University
\emails
   mbalducc@sju.edu,   
   michael.gelfond@ttu.edu,
\{epontell, tson\}@cs.nmsu.edu 
} %\\

@@

\begin{lstlisting}[language=clingo,numbers=none,caption=Set $RD^s$ of statements for the self-defense example, label=lst:pi_S, mathescape=true,xleftmargin=.01\textwidth, breaklines=true]
causes(appears(X),is(X,present),[]).
default($d^s_1$,inImminentDanger,[has(att,k),
  is(att,present),is(att,menc)])
default($d^s_2$,$\neg$inImminentDanger,[has(att,k),
  is(att,present),is(att,menc),is(k,fake)])
prefer($d^s_2$,$d^s_1$, [ ]).
default($d^s_3$,$\neg$inImminentDanger,
  [at(a1), is(a1,good)])
prefer($d^s_1$,$d^s_3$, [ ]).
\end{lstlisting}
The representation uses fluents $has(p,o)$ (p has/holds object o), $is(o,a)$ (object o has attribute a), and $at(l)$ (Bob is at location $l$). The objects are $k$ (the knife), $a_1$ (the good area of the city), and $att$ (the attacker). Additionally, we attributes constants $present$ (for attribute ``physically present''), $fake$ (attribute of the knife), $menc$ (menacing-looking), and $good$ (attribute of the area of the city). The actions are $appears(p)$ (person $p$ appears) and $shoots(p)$ (Bob shoots $p$). The effect of the former is described by the $causes$ statement; the latter action has no effects within the scope of the example. The intuitive meaning of the defaults is:
\begin{itemize}
\item $d^s_1$: If there is someone who is menacing-looking and holding a knife, one will normally believe to be in imminent danger
\item $d^s_2$: In the above case, if the knife is fake, one will not normally believe to be in imminent danger
\item $d^s_3$: One does not normally believe to be in imminent danger in a good area of the city
\end{itemize} 
As can be seen from the $prefer$ statements, $d^s_2$ is the most preferred default and $d^s_3$ the least preferred.
The initial part of the story is captured by:
\[
\begin{array}{lll}
O^s_{init}  \!=\!  \left\{
\begin{array}{ll} 
\!\!obs(at(a_1),0).   %\ \ \ \ 
&\!\! obs(is(a_1,good),0). \\
\!\!obs(\neg is(att,present),0). & \!\!
occ(appears(att),0). \\    % the guy appears
\!\!obs(is(att,menc),1). &\!\!  %\ \ \ \     % the guy is menacing-looking
obs(has(att,k),1).
\end{array}
\right.
\end{array}
\]
Let us consider what our framework yields for $KB^s_{init}=\langle O^s_{init},RD^s \rangle$. One can check that the answer set of $\Pi(KB^s_{init})$ contains $h(at(a_1),0)$ and $h(is(a_1,good),0)$. The only applicable default is $d^s_3$, yielding $believes(\neg inImminentDanger,0)$, i.e. Bob does not believe to be in imminent danger at time step $0$. The two possible evolutions of the story are formalized as follows:
\begin{itemize}
\item Scenario 1: \[
\begin{array}{lll}
O^s_{1} & = & \left\{
\begin{array}{l} 
occ(shoots(bob,att),1). 
\end{array}
\right\}\\
\ \\
\multicolumn{3}{l}{KB^s_1=\langle O^S_{init} \cup O^s_1,RD^s \rangle}
\end{array}
\]
The time step of interest is $1$. Note that $h(is(att,menc),1)$ and $h(has(att,k),1)$ follow trivially from the observations. Defaults $d^s_1$ and $d^s_3$ are thus applicable, while $d^{s}_2$ is not.  Because $d^s_1$ is preferred to $d^s_3$, the answer set of $\Pi(KB^s_1)$ contains $believes(inImminentDanger, 1)$, meaning that, at that time Bob pulled the trigger, he believed to be in imminent danger at the time of the shooting. It is worth noting how the framework is able to capture the evolution of Bob's beliefs over time.  
\item Scenario 2: \[
\begin{array}{lll}
O^s_{2} & = & \left\{
\begin{array}{l} 
obs(is(k,fake),1). \\   % the knife is fake
occ(shoots(bob,att),1).
\end{array}
\right\} \\
\ \\
\multicolumn{3}{l}{KB^s_2=\langle O^S_{init} \cup O^s_2,RD^s \rangle}
\end{array}
\]
At time step $1$, Bob has observed that the knife is fake and, thus, $h(is(k,fake),1)$ holds. Hence, all defaults are applicable.  Because $d^s_2$ is preferred to $d^s_1$, the answer set of $\Pi(KB^s_2)$ contains $believes(\neg inImminentDanger, 1)$. That is, Bob did not initially believe to be in imminent danger and did not believe so at the time of the shooting either. 
\end{itemize}

@@

%
\begin{figwindow}[3,r,%
        {\includegraphics[scale=0.44]{mim}},%
        {MITM attack carried out by Stuxnet: safe commands sent by the PLC (green) are replaced by dangerous ones (red); alarm readings from the centrifuge's sensors (red) are replaced by  safe ones\label{fig:stuxnet}}]
\end{figwindow}
The behavior of Stuxnet's MITM component is outlined in Figure \ref{fig:stuxnet}. Its component operates by intercepting the commands sent by a PLC to a connected centrifuge. The malware first increased the speed of the centrifuge above normal levels for a short amount of time, and later slowed it down below normal levels for a longer period of time. It is believed that the resulting stress caused components of the centrifuge to expand and eventually destroy it. Under normal conditions, sensors installed in the centrifuge would have alerted the PLC---and its users---about the abnormal conditions, giving them a chance to shut down the system before damage occurred. However, as part of the MITM attack, Stuxnet also intercepted the sensor readings from the centrifuge, and sent to the PLC fake readings based on previous recordings that indicated that the system was operating normally (this is known as a \emph{replay attack}).

@@

The relevant information can be formalized as shown below. The signature is defined so that $alert$ and $winter$ are fluents, while all others are beliefs. %using the following statements:
\iffalse
{
\[
\begin{array}{l}
default(d_1^m, hot\_room, [alert]). \\
default(d_2^m, \neg hot\_room, [cold\_weather]). \\
prefer(d_2^m,d_1^m, [ \: ]).\\
rule(r_1^m,cold\_weather,[winter]). \\
default(d_3^m,overheat,[alert, \neg hot\_room]).
\end{array}
\]
}
\fi
%
\begin{lstlisting}[language=clingo,numbers=none,caption=Relevant Information, label=lst:pi_M, mathescape=true,xleftmargin=.01\textwidth, breaklines=true]
default($d_1^m$, hot_room, [alert]). 
default($d_2^m$, $\neg$hot_room, [cold_weather]). 
prefer($d_2^m$,$d_1^m$, [ ]). 
rule($r_1^m$,cold_weather,[winter]). 
default($d_3^m$,overheat,[alert, $\neg$hot_room]).
\end{lstlisting}

@@

\section*{Acknowledgments}
